---
title: "Lab 04 - La Quinta is Spanish for next to Denny's, Pt. 1"
author: "John Bennett"
date: "Feb 20, 2022"
output: github_document
---

### Load packages and data

Note/Issue: I have inserted an "x" prior to the code chunk below in order to allow for knitting:

x```{r load-packages1}
install.packages("devtools")
devtools::install_github("rstudio-education/dsbox")
```

```{r load-packages2}
library(tidyverse) 
library(dsbox) 
```

```{r load-data, message = FALSE}
states <- read_csv("data/states.csv")
```

### Exercise 1
What are the dimensions of the Denny’s dataset? (Hint: Use inline R code and functions like nrow and ncol to compose your answer.) 
1643 rows and 6 columns

What does each row in the dataset represent? A fine dining location
What are the variables? Address of the location, City, State, zip, latitude, longitude


```{r dimensions}
nrow(dennys)
ncol(dennys)
spec(dennys)
```


### Exercise 2
What are the dimensions of the La Quinta’s dataset? 909 rows, 6 columns
What does each row in the dataset represent? An inn location
What are the variables? Address of the location, City, State, zip, latitude, longitude

```{r dimensionsLq}
nrow(laquinta)
ncol(laquinta)
spec(laquinta)
```

### Exercise 3
Are there any La Quinta’s locations outside of the US? If so, which countries? 
There are LaQuintas outside of the US, which were easy to find using their search filters. There are locations in Turkey, for example. 

What about Denny’s?
On the website, the offer a list of locations but all appear to be located in US states.


### Exercise 4
I would try the "list_unique <- lapply" function. This will give me a list of all of the values that fall under each column variable. 

### Exercise 5
I expanded the "dn" in the recommended code to spell "dennys".
With that, there do not appear to be any Dennys locations outside of the US, based on the results below:

```{r Ex5}
dennys %>%
  filter(!(state %in% states$abbreviation))
```
    
    
I tried this in the event that I have the code above wrong:

```{r Ex5a}
dennys %>% count(state, sort = TRUE)
```

These seemed potentially helpful, depending on what we go on to do with the data
```{r StateGoodies}
state.name
state.abb
```


    Find the Denny’s locations that are outside the US, if any. To do so, filter the Denny’s locations for observations where state is not in states$abbreviation. The code for this is given below. Note that the %in% operator matches the states listed in the state variable to those listed in states$abbreviation. The ! operator means not. Are there any Denny’s locations outside the US?

    Hint: Some of the abbreviations may not be familiar to you. Professor Google might be able to help.

    “Filter for states that are not in states$abbreviation.”

dn %>%
  filter(!(state %in% states$abbreviation))
  
### Exercise 6

```{r Ex6}
dennys %>%
  mutate(country = "United States")
```


    Add a country variable to the Denny’s dataset and set all observations equal to "United States". Remember, you can use the mutate function for adding a variable. Make sure to save the result of this as dn again so that the stored data frame contains the new variable going forward.

    Comment: We don’t need to tell R how many times to repeat the character string “United States” to fill in the data for all observations, R takes care of that automatically.

### Exercise 7 
Canada: British Columbia, Ontario
Mexico:  Aguascalientes(AG), Ciudad Juarez(CH), Poza Rica(VE), Puebla(PU), San Jose Chiapa(PU), San Luis Potosi(SL)
China: Weifang
New Zealand: Queenstown
Central America:  Honduras(FM)
Turkey: Bodrum, Giresun, Istanbul
United Arab Emirates: Dubai
Chile: Santiago
Colombia: Medellin


### Exercise 8 

```{r Ex8}
laquinta %>%
  mutate(country = case_when(
    state %in% state.abb     ~ "United States",
    state %in% c("ON", "BC") ~ "Canada",
    state == "ANT"           ~ "Colombia",
    state %in% c("AG",  "CH", "SL", "VE", "PU")          ~ "Mexico",
    state == "FM"           ~ "Honduras",
  ))
```


Going forward we will work with the data from the United States only. All Denny’s locations are in the United States, so we don’t need to worry about them. However we do need to filter the La Quinta dataset for locations in United States.


Issue: I'm getting an error here that "object 'country' not found" - this is erroneous!
I have inserted an "x" prior to the code chunk in order to allow for knitting

x```{r Ex8.1}
laquinta <- laquinta %>%
filter(country == "United States")
```


### Exercise 9 

```{r Ex9}
dennys %>%
  count(state) %>%
  inner_join(states, by = c("state" = "abbreviation"))
```

    Which states have the most and fewest Denny’s locations? What about La Quinta? Is this surprising? Why or why not?

Next, let’s calculate which states have the most Denny’s locations per thousand square miles. This requires joinining information from the frequency tables you created in the previous set with information from the states data frame.

First, we count how many observations are in each state, which will give us a data frame with two variables: state and n. Then, we join this data frame with the states data frame. However note that the variables in the states data frame that has the two-letter abbreviations is called abbreviation. So when we’re joining the two data frames we specify that the state variable from the Denny’s data should be matched by the abbreviation variable from the states data:

dn %>%
  count(state) %>%
  inner_join(states, by = c("state" = "abbreviation"))

Before you move on the the next question, run the code above and take a look at the output. In the next exercise, you will need to build on this pipe.

### Exercise 10 

```{r Ex10.1}
dennys <- dennys %>%
  mutate(establishment = "Denny's")
laquinta <- laquinta %>%
  mutate(establishment = "La Quinta")
```
    Which states have the most Denny’s locations per thousand square miles? What about La Quinta?

Next, we put the two datasets together into a single data frame. However before we do so, we need to add an identifier variable. We’ll call this establishment and set the value to "Denny's" and "La Quinta" for the dn and lq data frames, respectively.

dn <- dn %>%
  mutate(establishment = "Denny's")
lq <- lq %>%
  mutate(establishment = "La Quinta")

Because the two data frames have the same columns, we can easily bind them with the bind_rows function:

```{r Bindrows}
dennys_laquinta <- bind_rows(dennys, laquinta)
```

dn_lq <- bind_rows(dn, lq)

We can plot the locations of the two establishments using a scatter plot, and color the points by the establishment type. Note that the latitude is plotted on the x-axis and the longitude on the y-axis.

```{r PlotLocations}
ggplot(dennys_laquinta, mapping = aes(x = longitude, y = latitude, color = establishment)) +
  geom_point()
```

ggplot(dn_lq, mapping = aes(x = longitude, y = latitude, color = establishment)) +
  geom_point()

The following two questions ask you to create visualizations. These vizualizations should follow best practices you learned in class, such as informative titles, axis labels, etc. See http://ggplot2.tidyverse.org/reference/labs.html for help with the syntax. You can also choose different themes to change the overall look of your plots, see http://ggplot2.tidyverse.org/reference/ggtheme.html for help with these.

### Exercise 11 

```{r Ex11}
ggplot(dennys_laquinta %>% filter(state == "NC"),
       mapping = aes(x = longitude, y = latitude, color = establishment)) +
  geom_point(alpha = 0.5)
```

Filter the data for observations in North Carolina only, and recreate the plot. You should also adjust the transparency of the points, by setting the alpha level, so that it’s easier to see the overplotted ones. Visually, does Mitch Hedberg’s joke appear to hold here?

### Exercise 12


```{r Ex12}
ggplot(dennys_laquinta %>% filter(state == "TX"),
       mapping = aes(x = longitude, y = latitude, color = establishment)) +
  geom_point(alpha = 0.5)
```

Now filter the data for observations in Texas only, and recreate the plot, with an appropriate alpha level. Visually, does Mitch Hedberg’s joke appear to hold here?